African Wild Dog Dispersal and Connectivity under Climate Change - Lessons Learned from Seasonal Flood Extremes

### 2023-03-14 Chapter_8/04_Manuscript/Manuscript.tex
various updates

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{Flooding of the Okavango Delta influences Connectivity for Dispersing
African Wild Dogs}

% List of Authors
\author{
  David D. Hofmann\textsuperscript{1,2,\S} \orcid{0000-0003-3477-4365} \and
  Dominik M. Behr\textsuperscript{1,2} \orcid{0000-0001-7378-8538} \and
  John W. McNutt\textsuperscript{2} \and
  Arpat Ozgul\textsuperscript{1} \orcid{0000-0001-7477-2642} \and
  Gabriele Cozzi\textsuperscript{1,2} \orcid{0000-0002-1744-1940}
}

@@

\textsuperscript{2} Botswana Predator Conservation Program, Private Bag 13,
Maun, Botswana.

@@

\textbf{Running Title:} Seasonal Flooding of the Okavango Delta and its
Consequences for African Wild Dog Dispersal and Connectivity

@@

\textbf{Keywords:} movement, connectivity, seasonality, dispersal, conservation,
permeability

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Climate change is expected to profoundly impact the life history of wild-living
animal populations. While the impact of climate change on the demographics of
local subpopulations has been studied repeatedly, little is known about the
consequences of environmental change on dispersal and connectivity.

We capitalize on a ``natural experimental setup'', the flood-pulse driven
environmental change across the Okavango Delta in northern Botswana, to
investigate the impact of changing environmental conditions on dispersal
patterns and connectivity of the endangered African wild dog (\textit{Lycaon
pictus}). For this, we simulate dispersal trajectories across the landscape of
the Okavango Delta under two extreme environmental scenarios; one assuming a
maximum flood extent, one assuming a minimum flood extent.

During maximum flood, we observe a reduction in connectivity and an increase in
dispersal durations between distinct habitat patches. Notably, dispersal into
the center of the Okavango delta is reduced by 80\% during maximum flooding. At
minimum flooding, conversely, the delta reveals vital dispersal corridors and
increases chances of successful dispersal into neighboring habitat patches by xx
percent. Concomitantly, dispersal durations to move between patches is reduced,
on average by xx days,at low flood levels.

Climate change is expected to critically impact the flood-pulsing cycles of the
Okavango delta and may therefore substantially alter landscape connectivity in
this area. Our analysis suggests that anticipating such alterations in
environmental conditions is key when implementing conservation strategies.
Besides a better understanding of the conservation needs for the African wild
dog, our study also presents a first step towards incorporating environmental
changes due to seasonality or climate change in dispersal and connectivity
analyses and demonstrates the usefulness of individual-based dispersal
simulations as a pertinent conservation tool to study impacts of environmental
change on dispersal success.

@@

\section{Introduction}
Climate change is expected to profoundly impact ecosystems worldwide with
far-reaching consequences for species living therein \citep{Ozgul.2010,
Radchuk.2019, IPCC.2022}. By altering environmental conditions, climate change
affects animal behavior \citep{Fuller.2016}, resource availability
\citep{Durant.2007}, population dynamics \citep{Paniw.2021}, and the
distribution of wild living animal populations \citep{Thomas.2004,
Thuiller.2006}. One potential life-history pathway through which species may
mediate the effects of environmental change is dispersal \citep{Anderson.2012}.
By bringing individuals away from their natal location \citep{Clobert.2012},
dispersal equips species with a means to adapt to climate change by tracking
favorable habitat conditions \citep{Raia.2012} and to shift into a different
part of their fundamental niches \citep{Kokko.2022}. Dispersal furthermore
provides resilience against changing conditions by ensuring genetic diversity
\citep{Frankham.2002, Leigh.2012, Baguette.2013}, facilitating the colonization
empty habitats \citep{Gustafson.1996, Hanski.1999b, MacArthur.2001}, and
promoting the reinforcement of weakened and small subpopulations
\citep{Brown.1977}. As such, it is little surpsing that improving connectivity
is viewed as the number one strategy to combat the adverse effects of climate
change \citep{Heller.2009}.

While dispersal offers a means to offset the demographic consequences of climate
change \citep{Kokko.2006, Hodgson.2009, Travis.2013}, it itself is a function of
climatic and environmental conditions (e.g. \citep{Behr.2020}). The link between
dispersal and climaticor environmental can either be indirect, for example if
the propensity of individuals to disperse depends on environmental conditions,
or direct, when the biophysical environment through which dispersers move
affects dispersal prospects \citep{Travis.2013}. As the latter case indicates,
dispersal is also inextricably linked to the concept landscape connectivity
\citep{Baguette.2013}, which is understood as the degree by which the landscape
facilitates or impedes movements \citep{Taylor.1993}. Only if a sufficient
degree of landscape connectivity is maintained, successful dispersal of
individuals can be guaranteed \citep{Fahrig.2003}. Unfortunately, our
understanding how changing environmental conditions influence connectivity and
dispersal is limited \citep{Kokko.2006, Travis.2013}. Because collecting
information on dispersing animals is difficult and because it is difficult to
assess how changing environmental conditions affect connectivity...

The continued degradation and destruction habitats across the globe impedes
dispersal, promotes inbreeding and genetic drift and increases the
likelihood of local extirpation due to stochastic events \citep{Sawyer.2011}.
Combined, these factors limit species ability to adapt to changing environmental
conditions \citep{Fahrig.2003}.

\cite{Ashrafzadeh.2019} found that cimate change is likely to substantially
reduce connectivity for mountain newts in Iran. Similarly, \cite{Luo.2021}
predicted significant reductions in connectivity for the giant spiny frog in
South-East Asia.

However, failing to take into account potential changes in climatic and
environmental conditions imperils our ability to design cost-effective
conservation strategies.

Here, we used naturally occuring seasonal flood-cycles to approximate the likely
changes in connectivity that will arise under amplified climatic conditions due
to climate change. While a better understanding of the ODs flooding regime under global change is of major interest, reliable predictions remain challenging due to the complex interplay between climate,
anthropogenic water use, animal behavior, and topography of the region.

Climate change
could lead to the disruption of existing dispersal corridors and the emergence
of new dispersal barriers but may also provide possibilities to access
previously unsuitable dispersal routes. Designing and implementing effective
conservation strategies that aim to improve landscape connectivity, such as
habitat restoration or the creation of wildlife corridors, requires better
understanding of such dynamics.

Albeit the impacts of climate change on connectivity cannot be directly
estimated, one can study how changing environmental conditions due to
seasonality influence patterns of dispersal and connectivity.

Habitat degradation and fragmentation lead to isolated subpopulations that have
a high associated risk of extinction due to stochastic events and generally
decreased genetic variability \citep{Melbourne.2008}.

More recently, individual-based movement models (IBMMs) have been uprising and
are recommended for their ability to incorporate several crucial realities of
animal behavior, such as, for example, autocorrelation, that are by omitted by
necessity using resistance-based approaches \citep{Hofmann.2023,
Unnithan.2022a}. IBMMs allow much greater flexibility and are thus considered
powerful tools for examining connectivity under different landscape
configurations \citep{Littlefield.2019}.

Botswana is among the countries most heavily influenced by climate change .

For multispecies approaches to connectivity within the KAZA-TFCA, see
\cite{Brennan.2020} and \cite{Lines.2021}.

Talk about connectivity models
Talk about functional connectivity
Talk about habitat fragmentation and destruction
Say that wild dogs are keystone predators and an umberella species for
conservation

Unfortunately, little is known about the specific responses of different species
to climate change, especially for regions most vulnerable to altered conditions
\citep{Paniw.2021}.

Earlier studies demonstrating the importance of rendering seasonal change
include \cite{Chetkiewicz.2009}, who use resource selection functions to demonstrate
that connectivity differs for Ursus actor and Puma concolor and \cite{Mui.2017},
who showed that connectivity for turtles changes depending on the season, and
\cite{Kaszta.2021}, who show monthly differences in elephant's habitat selection
using path-selection functions.

Only if there

The challenges associated with studying dispersing animals have thus far
impaired the collection of data of dispersers at the appropriate temporal and
spatial scale \citep{Graves.2014, Vasudev.2015} and weakened our ability to
project dispersal prospects under changing environmental conditions into the
future. Nevertheless, a more comprehensive understanding of dispersal under
changing environmental conditions is critical for designing effective
conservation strategies in the light of climate change. By predicting and
modeling dispersal patterns under different scenarios, we can better anticipate
the impacts of climate change on species and design effective conservation
strategies to protect them. This is especially important for endangered species
that are already at the verge of extinction.

The Okavango delta in Southern Africa poses a unique opportunity to study the
impacts of environmental change on species dispersal ability and connectivity in a
large scale natural experiment setup. According to projections by the IPCC,
Botswana and its surroundings are among the most vulnerable to climate change.
While global temperatures are expected to increase by xx degrees until the end
of the 21st century, For example, \cite{Engelbrecht.2015} predicts an increase
between 4 and 6\degree C for regions in southern Africa. One species that
heavily relies on dispersal for long-term persistence and that inhabits such an
arid ecosystem in Southern Africa is the African wild dog (\textit{Lycaon
pictus}). The African wild dog is a lightly built canid lives in cohesive packs
comprising up to 20 individuals. Wild dogs are social breeders and the majority
of reproduction is monopolized by the dominant alpha couple who are supported by
subordinates in taking care of their pups. The species exhibits strong
inbreeding avoidance, hence juvenile individuals born into the pack disperse in
same-sex sibling coalitions after reaching sexual maturity \citep{McNutt.1996,
Behr.2020}. Notably, the African wild dog is the sole extant representative of
its genus and thus considered as high-priority species for conservation efforts
\citep{Leigh.2012}. However, fragmentation and destruction of their historic
habitat may have decreased dispersal success for this species and thus
jeopardizes connectivity \citep{Leigh.2012}. Since the impacts of climate change
are not uniformly distributed across the planet, but particularly pronounced at
high latitudes and arid ecosystems, the species may be seen as most vulnerable
to such ecosystem shifts.

Talk about Zeller und Osipova

In a previous study, \citep{Hofmann.2022} have investigated landscape
connectivity in the same ecosystem, yet assuming a static landscape.

Here, we capitalize on a previously parameterized dispersal model and use
dispersal simulations to investigate dispersal success and connectivity patterns
for African wild dogs under two extreme scenarios; one assuming maximum flooding
of the Okavango delta and one assuming minimum flooding of the Okavango delta.
Given that water poses a barrier to dispersing individuals, we anticipated that
dispersal prospects and connectivity during maximum flood are low. During
minimum flood, in contrast, we expect to reveal the presence of several
dispersal corridors that may be used to freely and quickly move between
remaining habitat patches.

\section{Materials and Methods}
We conducted all analyses using the programming language \textsf{R}. Any spatial
data manipulation was completed using the \textsf{terra} and \textsf{spatstat}
packages. Several helper functions for the simulation algorithm were written in
\textsf{C++} and imported to R using the \textsf{Rcpp} package. Figures were
generated using \textsf{ggplot2}, \textsf{igraph}, and \textsf{ggnetwork}. All
R-scripts used to conduct our analyses are provided through an online
repository.

\subsection{Study Area}
The study area for this analysis was focused on the Okavango delta (OD) and its
surroundings in Southern Africa, comprising parts of Angola, Namibia, Botswana,
Zimbabwe, and Zambia (\Cref{StudyArea}). The OD is the world's largest inland
delta and the main driver of seasonal environmental change in the region. While
our primary focus lied on the immediate surroundings of the Okavango Delta, we
considered a large rectangular extent stretching from 20\degree 30' E to
26\degree E. (totaling to an area of 300'000 km\textsuperscript{2}) to
accommodate for the long distance dispersal events commonly observed in African
wild dogs (e.g. \citealp{Davies-Mostert.2012, Masenga.2016, Cozzi.2020}). The
flood-pulsing rhythm of the OD is mainly dictated by precipitation in the
catchment areas of the Angolian highlands, where rainwater is collected and
channeled into the OD through the Okavango River. Although precipitation in
Angola peaks between December and March, water only slowly descends through the
Okavango river and its distributaries, reaching the distal ends of the delta in
July or August, where the water percolates at the Thamalakane and Kunyere
Faults. At minimum extent, the flood covers an area of 3'600
km\textsuperscript{2}, during maximum flood more than 9'000
km\textsuperscript{2}. Vegetation in the study area is dominated by mopane
forest, mixed acacia woodland, and grassland. Human influence is low and mainly
concentrated around small villages at the western periphery of the delta as well
as the city of Maun at the south-eastern tip of the OD. Large portions of land
are dedicated national parks, game reserves or forest reserves. The study area
is also part of the world's largest transboundary conservation initiative, the
Kavango-Zambezi Transfrontier Conservation Area, which aims to restore
connectivity between protected areas in Southern Africa.

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_StudyArea.png}
  \caption{Study area across which we simulated dispersal. Simulated dispersers
  were released at random locations within the orange source areas distributed
  across the delta. Emigration zones (purple) served as checkpoints and enabled
  us to identify if and where simulated dispersers left the close surroundings
  of the Okavango delta. These zones were generated using a set of cutlines
  originating from the center of the delta and roughly cutting an elliptical
  buffer zone into sections of equal size.}
  \label{StudyArea}
  \end{center}
\end{figure}

\subsection{Spatial Habitat Layers}
We represented the physical landscape through which dispersers could move by a
set of spatially referenced habitat layers, each resolved at 250m x 250m. The
set of layers included water-cover, distance-to-water, tree-cover,
shrub/grassland-cover, and a human influence layer depicting anthropogenic
influences through villages, roads, and agriculture. A detailed description of
the different habitat layers is provided in \Cref{Hofmann.2021}. Importantly,
the water-cover and derived distance-to water layers were generated using MODIS
Terra MCD43A4 satellite imagery that was classified using a ``floodmapping''
algorithm developed by \citep{Wolski.2017} and available through the R-package
\textsf{floodmapr}. The algorithm allowed us to generate almost weekly updated
``floodmaps'', thus providing detailed information about the flood-extent at any
given point in time. In total, we generated 700 floodmaps between the years 2000
and 2019. Based on these maps, we generated a minimal and maximum flood
scenario. To create the minimum flood scenario, we averaged the 50 floodmaps
with smallest flood extent and generated a binary image using areas that were
inundated in at least 50\% of the maps. Similarly, we created an average image
for high flood using the 50 most flooded maps. The final maps are depicted in
\Cref{FloodExtent}.

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_FloodExtent.png}
  \caption{Flood extent in the two scenarios considered. In the left panel, the
  flood is at an extremely low level, stretching across 3'558
  km\textsuperscript{2}, whereas in the right panel the flood is at an extremely
  high level and covers 9'397 km\textsuperscript{2}. The two maps were generated
  using 700 remote sensed MODIS MCD43A4 satellite images spanning the years 2000
  to 2019.}
  \label{FloodExtent}
  \end{center}
\end{figure}

\subsection{Dispersal Model}
Our dispersal model was based on a previously parametrized and validated
integrated step-selection function (iSSF, \citealp{Avgar.2016}) fitted to GPS
data of dispersing AWDs \citep{Hofmann.2021}. In step selection functions (SSFs,
\citealp{Fortin.2005}), observed GPS locations are converted into steps (the
straight-line traveled between two GPS recordings \citep{Turchin.1998a}) and
compared to a set of \textit{random} steps in a (mixed effects) conditional
logistic regression framework \citep{Fortin.2005, Thurfjell.2014, Muff.2020,
Fieberg.2021}. The model presented in \citep{Hofmann.2021} used dispersal data
of 16 dispersing African wild dogs from a free-ranging wild dog population in
northern Botswana. GPS data during dispersal was collected at 4-hourly intervals
and translated into steps of similar duration. Observed steps were then paired
step with 24 random steps that were generated using a uniform distribution for
turning angles (\(-\pi, +\pi\)) and step lengths from a gamma distribution
fitted to observed steps (scale \(\theta\) = 6'308 and shape \(k\) = 0.37). It
was then assumed that animals assigned to each observed and random step a
selection score of the form \citep{Fortin.2005}:

\begin{equation}
\label{EQ1}
  w(x) = exp(\beta_1 x_1 + \beta_2 x_2 + ... + \beta_n x_n)
\end{equation}

Where (\(x_1, x_2, ..., x_n\)) represent the covariate values along each of the
steps and the (\(\beta_1, \beta_2, ..., \beta_n\)) are the animal's relative
selection strengths \citealp{Avgar.2017} towards these covariates. The benefit
of \textit{integrated} SSFs over regular SSFs is that they provide a means to
render two complementary ``kernels''. A movement kernel that describes general
movement behavior of dispersing AWDs and a habitat kernel that describes
preferences of AWDs with regards to environmental conditions
\citep{Fieberg.2021}. iSSFs also allow interactions among the two kernels and
are thus suitable to render that movement behavior may change depending on
habitat conditions. A brief summary of the fitted dispersal model is provided in
Appendix xx.

\subsection{Source Areas and Emigration Zones}
We simulated dispersing AWDs originating from six distinct source areas located
in the vicinity of the Okavango delta (\Cref{StudyArea}). For areas one to six
we selected locations at the delta's periphery that remained dry in both
scenarios. At those locations, we generated circular buffers with a radius of 20
km. For source area six, we isolated a polygon covering Chief's Island, a
peninsula located at the OD's center. Besides source areas, we also generated
``zones of emigration'' that we used as checkpoints to determine if and where
simulated individuals left the delta's vicinity (\Cref{StudyArea}). We generated
emigration zones by first overlaying the OD with an elliptic buffer zone that we
dissected using a set of cutlines that originated from the ODs center and spread
according to cardinal points (\Cref{StudyArea}).

>>>>

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
% \title{Flooding of the Okavango Delta influences Connectivity for Dispersing
% African Wild Dogs}
\title{African Wild Dog Dispersal and Connectivity under Climate Change -
Lessons Learned from Seasonal Flood Extremes}

% List of Authors
\author{
  David D. Hofmann\textsuperscript{1,2,\S} \orcid{0000-0003-3477-4365} \and
  Dominik M. Behr\textsuperscript{1,2} \orcid{0000-0001-7378-8538} \and
  John W. McNutt\textsuperscript{2} \and
  Arpat Ozgul\textsuperscript{1,2} \orcid{0000-0001-7477-2642} \and
  Gabriele Cozzi\textsuperscript{1,2} \orcid{0000-0002-1744-1940}
}

@@

\textsuperscript{2} Botswana Predator Conservation Program, Wild Entrust,
Private Bag 13, Maun, Botswana.

@@

\textbf{Running Title:} African Wild Dog Dispersal in a Changing Climate:
Lessons Learned from Seasonal Flood Extremes

@@

\textbf{Keywords:} movement, connectivity, climate change, seasonality,
dispersal, conservation, individual-based simulations, step-selection

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Climate change is expected to profoundly impact the life history of wild-living
animal populations. While the impacts of climate change on the demographics of
local subpopulations have been studied repeatedly, little is known about the
consequences for dispersal and connectivity.

We capitalize on a ``natural experimental setup'', the flood-pulse driven change
in surface-water across the Okavango Delta in northern Botswana, to investigate
the impact of changing environmental conditions on dispersal patterns and
connectivity of the endangered African wild dog (\textit{Lycaon pictus}). For
this, we simulate dispersal trajectories across the Okavango Delta under two
extreme scenarios that serve to represent environmental conditions akin to those
expected under continued climate change; one assuming a maximum flood, one
assuming a minimum flood.

During maximum flood, the Okavango Delta poses an important dispersal barrier
that reduces dispersal prospects in increases dispersal durations between
distinct areas. Across the entire study area, we observe
$\input{99_NumberReachingOthersPercentage.tex}$\% lower dispersal success and
$\input{99_StepsToReachingOthersPercentage.tex}$\% longer dispersal durations
during maximum flood. Most notably, dispersal into the central habitats of the
Okavango Delta is reduced by $\input{99_NumberReachingArea6Percentage.tex}$\%
with an accompanied increase in dispersal durations of
$\input{99_StepsToReachingArea6Percentage.tex}$\%. Depending on the flood,
dispersal corridors traversed different areas and dispersers moved into
proximity of different human-dominated areas.

Whilst the exact impacts of climate change on the flooding regime of the
Okavango Delta remain unknown, our results suggest that connectivity will vastly
differ depending on future flood conditions. Acknowledging such differences will
be key to design effective conservation strategies, especially in light of
ongoing climate change. Since we highlight critical dispersal corridors and
human-wildlife conflict zones for two distinct future scenarios, our results
will facilitate the evidence-based conservation of the endangered African wild
dog.

@@

\section{Introduction}
\subsection{Climate Change and Dispersal}
Climate change is expected to profoundly impact ecosystems across the globe with
far-reaching consequences for the species living therein \citep{Ozgul.2010,
 Radchuk.2019, IPCC.2022}. By altering environmental conditions, climate change
affects animal behavior \citep{Fuller.2016}, resource availability
\citep{Durant.2007}, population dynamics \citep{Paniw.2021}, and the
distribution of wild living animal populations \citep{Thomas.2004,
 Thuiller.2006}. An important life-history pathway through which species may
mediate the negative consequences of climate change is dispersal
\citep{Anderson.2012}, i.e. the movement of individuals away from their natal
location to the site of first reproduction \citep{Clobert.2012}. Through
dispersal, species may adapt to climate change by tracking favorable habitat
conditions \citep{Raia.2012} and by shifting into a different region of their
fundamental niche \citep{Kokko.2006}. Dispersal also facilitates the
colonization empty habitats \citep{Gustafson.1996, Hanski.1999a,
 MacArthur.2001}, promotes the reinforcement of weakened and small subpopulations
\citep{Brown.1977}, and safeguards genetic diversity \citep{Frankham.2002,
 Leigh.2012, Baguette.2013}, thus providing additional resilience against
changing environmental conditions \citep{Kokko.2006, Fahrig.2003}.

\subsection{Connectivity}
While dispersal offers a means to offset the negative demographic consequences
of climate change \citep{Kokko.2006, Hodgson.2009, Travis.2013}, it itself is a
function of climatic and environmental conditions (e.g. \citealp{Elliot.2014,
 Behr.2020}). The link between dispersal and the environment can either be
indirect, for example if the propensity of individuals to disperse depends on
environmental conditions, or direct, when the biophysical environment through
which dispersers move affects dispersal prospects \citep{Travis.2013}. The
latter highlights that dispersal is also inextricably linked to the concept
landscape connectivity \citep{Baguette.2013}, which is understood as the degree
to which the landscape facilitates or impedes movements \citep{Taylor.1993}. A
sufficient degree of landscape connectivity is thus a critical prerequisite for
successful dispersal \citep{Fahrig.2003}, yet the continued degradation and
destruction habitats worldwide continues to imperil the dispersal prospects of
many species \citep{Melbourne.2008, Sawyer.2011}. Conservation strategies that
aim at facilitating dispersal by improving landscape connectivity are therefore
often viewed as pinnacle of all conservation strategies \citep{Heller.2009}.
Despite this, our understanding of dispersal and its implications for
connectivity is limited, especially in light of changing environmental
conditions.

\subsection{Modeling Connectivity}
To study dispersal and connectivity, various modeling techniques have emerged
(see e.g. \citealp{Etherington.2016} and \citealp{Diniz.2019} for overviews).
Initially, the techniques were limited to examining structural aspects of
connectivity by focusing on the composition and configuration of habitat
patches, while ignoring species' responses to the landscape matrix
\citep{Tischendorf.2000, Doerr.2011}. With the increasing availability of
telemetry data and methods to study species' habitat and movement preferences
\citep{Boyce.2002, Fortin.2009, Cushman.2010, Avgar.2016}, preferably during
dispersal \citep{Elliot.2014}, however, the focus has shifted towards a more
functional view on connectivity, which also takes into account how species
interact with their surroundings \citep{Tischendorf.2000, Doerr.2011}.
Currently, the most prominent \textit{functional} connectivity models are based
on least-cost path analysis (LCPA, \citealp{Adriaensen.2003}) and circuit theory
(CT, \citealp{McRae.2008}), two graph-based methods that estimate conductance of
the landscape by means of a resistance (or inversely permeability) surface
\citep{Zeller.2012}. Such a surface is meant to reflect the ease or willingness
at which the focal species traverses a specific area and is generated by
consolidating multiple habitat layers into a single layer of resistance
\citep{Zeller.2012}. Since both LCPA and CT approaches make assumptions that are
rarely met by dispersing individuals, individual-based movement models (IBMMs),
in which dispersal movements are simulated explicitly, have also gained some
momentum \citep{Kanagaraj.2013, Allen.2016, Hauenstein.2019, Diniz.2019,
 Zeller.2020, UnnithanKumar.2022, UnnithanKumar.2022a, Hofmann.2023}. IBMMs
provide great modeling flexibility and are thus considered powerful tools for
examining connectivity under different landscape configurations
\citep{Littlefield.2019, UnnithanKumar.2022a}. However, most connectivity
studies focus on a snapshot in time and fail to account for changing
environmental conditions, such as those akin to climate change. Moreover, the
challenges associated with studying dispersing animals further impairs the
collection of data during dispersal at the appropriate temporal and spatial
scale \citep{Graves.2014, Vasudev.2015} and weakens our ability to project
dispersal prospects under changing environmental conditions into the future.

\subsection{Climate Change and Seasonality}
Predicting the impacts of climate change on dispersal and connectivity is
non-trivial and typically requires spatial information about future climatic or
environmental conditions over the area of interest \citep{Littlefield.2019}.
This information can then be used in various ways. \cite{Ashrafzadeh.2019}, for
example, combined climatic predictions until 2070 with a species-distribution
model for mountain newts (\textit{Neurergus kaiseri}) in Iran to demonstrate a
decrease in connectivity due to increased habitat fragmentation. Similarly,
\cite{Luo.2021} mapped the future distribution of the giant spiny frog
(\textit{Quasipaa spinosa}) under different representative climate pathways and
reported a reduction in connectivity for the species across South-East Asia. In
these studies, the focus lies on the impacts of climate change on species
distribution and subsequent changes in connectivity due to the configuration of
habitat patches, yet less on the habitat matrix and its implications for
dispersal. For martens, (\textit{Martes americana}), \cite{Wasserman.2012}
developed several resistance layers emerging under different climate scenarios
and find that already low warmings will result in increased isolation of
remaining subpopulations. While not primarily focused on climate change, another
body of literature captures environmental variability by generating resistance
surfaces for different scenarios. \cite{Mui.2017}, for instance, developed
seasonal resistance maps for Blanding's turtle \textit{Emydoidea blandingii}
showing that connectivity was substantially lower in late summer compared to
spring. Similarly, \cite{Osipova.2019} studied connectivity for African
elephants (\textit{Loxodonta africana}) during wet and dry season and found that
ignoring seasonality resulted in an underestimation of connectivity during the
wet season and an overestimation during the dry season. For the same species,
\cite{Kaszta.2021} provide monthly updated connectivity maps revealing that
connectivity varies strongly across a typical year. Finally, \cite{Zeller.2020}
use dynamic resistance surfaces showing differences in connectivity for black
bears \textit{Ursus americano}. Altogether, the studies exemplify that
connectivity should not be regarded as static in time, but dynamic across and
within years.

In many cases, anticipating environmental conditions under climate change is not
viable as relevant data is not available or entails major uncertainty
\citep{Collins.2012}. This is particularly true for complex ecosystems with
intricate feedback loops and in cases where one is interested in landscape
characteristics, rather than climatic conditions. In general, it is accepted
that aside from increasing temperatures, climate change will also amplify the
frequency and magnitude of extreme events, such as severe droughts, heavy
precipitation, floods, and storms \citep{Stott.2016, Ummenhofer.2017,
 IPCC.2022}. Thus, instead of attempting to study the impacts of climate change
directly, one may capitalize on naturally occurring fluctuations of the
environment to gauge the likely consequences of shifting the system towards what
is currently considered an extreme.

\subsection{Okavango Delta}
The Okavango Delta (OD) in Southern Africa poses a unique opportunity to study
the impacts of extreme environmental change on species dispersal ability and
connectivity in a large scale natural experiment setup. The OD is the world's
largest inland delta and characterized by substantial seasonal differences in
surface-water cover. Throughout the course of a year, the area covered by the
OD's floodwaters can fluctuate between 3'000 and 10'000 km\textsuperscript{2}
with striking variability within and between years \citep{Gumbricht.2004,
 Wolski.2017}. Importantly, the region is among the most vulnerable to climate
change, as a temperature increase of 4 to 6\degree C above pre-industrial levels
is expected within the 21\textsuperscript{st} century \citep{Engelbrecht.2015,
 Akinyemi.2019}, which is far beyond the global average \citep{IPCC.2022}. A
keystone predator in this ecosystem and an umbrella species for conservation
efforts is the African wild dog (AWD, \textit{Lycaon pictus}). While the species
was once widespread across across entire Sub-Saharan Africa, it has disappeared
from a vast majority of its historic range, mainly due to human persecution,
deadly diseases, and continued destruction and degradation of its habitats
\citep{Woodroffe.2012}. AWDs are characterized by an unsurpassed dispersal
ability, as young individuals that leave their natal pack can cover several
hundred kilometers within a few days \citep{Davies-Mostert.2012, Masenga.2016,
 Cozzi.2020}. Dispersal typically happens in dispersal coalitions of same-sex
siblings  \citep{McNutt.1996}. Previous research has shown that the ODs
floodwaters pose a major dispersal barrier, yet the analysis was centered around
an average flooding scenario \citep{Hofmann.2021}.

\subsection{What We Did}
Here, we utilize a previously parameterized and validated dispersal model as
IBMM to simulate dispersal and assess dispersal success and connectivity
patterns for African wild dogs under two extreme scenarios: one assuming maximum
flooding of the Okavango delta and one assuming minimum flooding of the Okavango
delta. The IBMM was trained using GPS data collected during dispersal and
frequently updated environmental data, thus providing a high degree of realism
\citep{Hofmann.2021}. Given that dispersers avoid crossing through water (albeit
we do occasionally observe it in the field), we anticipated that dispersal
prospects and connectivity during maximum flood would be low. Moreover,  when
the flood extent of the OD is at its maximum, the water extends almost into the
densely populated village of Maun. Since both the flood and humans are avoided
by dispersing wild dogs, we anticipated that a fully flooded Delta would result
in a total halt of movement between the Western and Eastern side of the OD.
Information on habitat selection or connectivity is also suitable for predicting
areas with an elevated potential for human wildlife conflict
\citep{Buchholtz.2020}. Hence, also quantified areas with high potential for
human wildlife conflict.

\section{Materials and Methods}
We conducted all data preparation and analyses using the programming language
\textsf{R} \citep{RCoreTeam.2022}. For any spatial data manipulation, we used
the packages \textsf{terra} \citep{Hijmans.2022} and \textsf{spatstat}
\citep{Baddeley.2015}. Several helper functions for the dispersal simulation
algorithm were written in \textsf{C++} and imported to R using the \textsf{Rcpp}
package \citep{Eddelbuettel.2011}. Network analysis was achieved in
\textsf{igraph} \citep{Csardi.2006} and figures were generated using
\textsf{ggplot2} \citep{Wickham.2016} and \textsf{ggnetwork}
\citep{Briatte.2021}. All R-scripts required to replicate our analyses are
provided through an online repository.

\subsection{Study Area}
The study area for this analysis was focused on the Okavango delta (OD) and its
surroundings in Southern Africa, comprising parts of Angola, Namibia, Botswana,
Zimbabwe, and Zambia (\Cref{StudyArea}). While our primary focus lied on the
immediate surroundings of the Okavango Delta, we considered an extended
rectangular extent stretching from 20\degree 30' E to 26\degree E. This
encompasses a total of 300'000 km\textsuperscript{2}) and comprises all long
distance dispersal events previously recorded in this area \citep{Cozzi.2020,
 Hofmann.2021}. The annual flood-pulsing rhythm of the OD is mainly dictated by
precipitation in the Angolan highlands, which serve as catchment areas from
which water is further channeled into the Okavango River and transported into
the OD \citep{Wolski.2017}. Although precipitation reaches its maximum between
December and March, the collected water only slowly descends through the
Okavango River and its distributaries, reaching the distal ends of the delta
only towards July or August. In result, peak flooding is out of sync with local
precipitation, such that the flood usually arrives in the OD during the peak dry
season. Once the water reaches the OD's distal ends, it percolates at the
Thamalakane and Kunyere Faults, two natural faultlines at which the waterflow is
hindered. During minimum extent, the flood covers an area of only 3'600
km\textsuperscript{2}, whereas during maximum flood more than 9'000
km\textsuperscript{2} are flooded. Vegetation in this study area is dominated by
mopane forest, mixed acacia woodland, and grassland. Human influence is low and
mainly concentrated around small villages at the western periphery of the OD as
well as the city of Maun at the south-eastern tip of the OD. Large portions of
land are dedicated national parks, game reserves or forest reserves. The study
area is also part of the world's largest transboundary conservation initiative,
the Kavango-Zambezi Transfrontier Conservation Area. Previous studies have
attributed a high potential of this initiative for improving connectivity for
various species \citep{Brennan.2020, Lines.2021, Hofmann.2021}.

\begin{figure}
 \begin{center}
  \includegraphics[width = \textwidth]{99_StudyArea.png}
  \caption{(a) Location of the study area in Southern Africa across which we
   simulated dispersing African wild dogs. To mitigate edge effects, our study
   area comprised a buffer zone (red polygon) within which we randomized
   covariate values of the habitat layers. (b) The study area was centered on the
   Okavango Delta and encompassed its immediate surroundings. We initiated
   simulated dispersers at random locations within one of the six source areas
   (orange polygons) that we distributed across the delta. Emigration zones
   (purple polygons) served as checkpoints to identify if and where simulated
   dispersers left the close surroundings of the Okavango delta. These zones were
   generated using a set of cutlines (black dotted lines) originating from the
   center of the delta that dissected an elliptical buffer surrounding the delta
   into sections of equal size and in accordance with the cardinal directions.}
  \label{StudyArea}
 \end{center}
\end{figure}

\subsection{Spatial Habitat Layers}
We represented the physical landscape through which dispersers could move by a
set of spatially referenced habitat layers, each with a resolution of 250 m. The
set of layers included water-cover, distance-to-water, tree-cover,
shrub/grassland-cover, and a human influence layer depicting anthropogenic
influences through villages, roads, and agriculture. A detailed description of
the different habitat layers is provided in \citep{Hofmann.2021}. Importantly,
the water-cover and derived distance-to water layers were generated using MODIS
Terra MCD43A4 satellite imagery that was classified into binary water-cover maps
using a ``floodmapping'' algorithm developed by \citep{Wolski.2017}. This
allowed us to generate almost weekly updated ``floodmaps'', thus providing
detailed information about the flood-extent at any given point in time. In
total, we generated 700 floodmaps between the years 2000 and 2019, based on
which we generated a minimal and maximum flood scenario. To create the minimum
(maximum) flood scenario, we averaged the 100 floodmaps with the smallest
(highest) flood extent and generated a binary layer by masking all pixels that
were inundated in less than 50\% of the maps. The resulting maps are presented
in \Cref{FloodExtent}. Ultimately, we combined the habitat layers into two
stacks, one representing the minimum flood scenario, one representing the
maximum flood scenario. To mitigate edge effects during the dispersal
simulation, we followed \cite{Koen.2010} and expanded the spatial extent of the
stacked layers by 20\% and randomized habitat values within the so created
buffer zone (red rectangle in \Cref{StudyArea}a).

\begin{figure}
 \begin{center}
  \includegraphics[width = \textwidth]{99_FloodExtent.png}
  \caption{Flood extent in the minimum (a) and maximum (b) flood scenarios. In
   the minimum flood scenario (a), water stretched across 3'558
   km\textsuperscript{2}, whereas during maximum flood (b) it covered 9'397
   km\textsuperscript{2}. The two maps were generated using 700 remote sensed
   MODIS MCD43A4 satellite images spanning the years 2000 to 2019.}
  \label{FloodExtent}
 \end{center}
\end{figure}
